---
title: "DSS Analysis"
author: "Cemile Balkas"
output: html_document
params:
  input1: "WT_WG.tidy.dssin.gz" # Reference group 
  input2: "3BKO_WG.tidy.dssin.gz" # Test group
---

To run on command line Rscript -e "rmarkdown::render('Analysis.Rmd', params = list(input1 = 'file_name', input2 = 'file_name'))"

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

This script performs analysis of differentially methylated regions (DMRs) using the DSS package. The input data is from Oxford Nanopore bisulfite sequencing and reformatted to make it compatible for DSS.

## Setup

```{r itit}
# Make sure the following packages are installed before running this document:
# Run this once in your R console (not inside this Rmd)
# BiocManager::install("DSS")

# Load required libraries
library(DSS)
require(bsseq)

```


```{r}
cat("Comparing methylation between:\n")
cat("  - Reference group:", params$input1, "\n")
cat("  - Test group:", params$input2, "\n")

```

# Read Files

```{r load-data}
# Load methylation count data: chr, pos, total reads (N), methylated reads (X)
wt_file <- params$input1
ko_file <- params$input2

wt <- read.table(wt_file, header = FALSE)
ko <- read.table(ko_file, header = FALSE)

colnames(wt) <- c("chr", "pos", "N", "X")
colnames(ko) <- c("chr", "pos", "N", "X")

# Create BSseq object
# This object holds methylation information in the structure DSS requires
BSobj <- makeBSseqData(list(wt, ko), c("WT", "KO"))

# View summary of the BSseq object
BSobj
```

## Box plot Coverage comparison between crhomsomomes

```{r coverage-boxplots}
# Canonical chromosome levels
canon_chrs <- paste0("chr", 1:22)

# Makes the chr column is a factor with correct order
wt$chr <- factor(wt$chr, levels = canon_chrs)
ko$chr <- factor(ko$chr, levels = canon_chrs)

# WT coverage boxplot
boxplot(N ~ chr, data = wt,
        main = "Coverage by Chromosome (WT)",
        ylab = "Coverage (N)",
        xlab = "Chromosome",
        col = "skyblue",
        outline = FALSE,
        las = 2, cex.axis = 0.7)

# KO coverage boxplot
boxplot(N ~ chr, data = ko,
        main = "Coverage by Chromosome (3BKO)",
        ylab = "Coverage (N)",
        xlab = "Chromosome",
        col = "coral",
        outline = FALSE,
        las = 2, cex.axis = 0.7)
```




##  Differential Methylation Loci (DML) test

DML test between the two groups (KO vs WT) is performed using the DMLtest function from the DSS package. This tests each CpG site for significant differences in methylation level between groups, applying smoothing to borrow strength from neighboring sites, but is also needed in this case because there are no replicates. 

```{r dml-test}
# Run DML test with smoothing (recommended for WGBS data and when there are no replicates)
dmlTest.sm <- DMLtest(BSobj,  group1="WT", group2="KO", smoothing=TRUE, ncores=4)
head(dmlTest.sm)
```
## Identify DMLs based on statistical test results

With the test results, the callDML function was called. The results DMLs are sorted by the significance.

```{r}
# Null hypothesis: the diff. in methylation levels is 0 between groups (mu1 = mu2)
# No minimum effect size required
dmls <- callDML(dmlTest.sm, p.threshold = 0.001)
head(dmls)

# Null hypothesis: the difference in methylation is less than or equal to delta (|mu1 - mu2| ≤ 0.1)
# Code below filters for biologically meaningful changes (at least 10%)
dmls_strict <- callDML(dmlTest.sm, delta = 0.1, p.threshold = 0.001)
head(dmls_strict)

# For DMLs
dmls_strict$start <- dmls$pos - 1
# dmls$end <- dmls$pos

```

## Identify DMRs by grouping nearby significant CpGs (DMLs)

After the identification of differentially methylated loci (DMLs), differentially methylated regions (DMRs) were identified. The callDMR() function from the DSS was used. Regions are identified based on significance thresholds for both statistical confidence (p-value) and the magnitude of methylation difference (delta). This step summarises regional methylation patterns than just the single loci differences.

```{r dmr-call}
# Detect DMRs based on smoothed DML test results
# p.threshold filters CpG sites with p-values below 0.01
dmrs = callDMR(dmlTest.sm, p.threshold=0.01)
head(dmrs)

# Apply an additional effect size threshold (e.g. delta = 0.1)
dmrs_strict = callDMR(dmlTest.sm, 
                      delta=0.1, 
                      p.threshold=0.01, 
                      minlen = 50, # min length of the DMR (def: 50bp)
                      minCG = 3, # min number of CpG sites required for DMR
                      dis.merge = 50, # when two DMRs are closer to each other than this value they are merged (def:50)
                      pct.sig = 0.9) # % of CG sites with significant p values
head(dmrs_strict)

showOneDMR(dmrs_strict[1,], BSobj)

dmrs$end <- dmrs$end + 2
dmrs_strict$end <- dmrs_strict$end + 2
head(dmrs)

# order dmrs using chromosome and start pos
# Set chromosome column as a factor with the correct order
dmrs_strict$chr <- factor(dmrs_strict$chr, levels = canon_chrs)

# Order by chr (with factor level order) and then by start
dmrs_strict <- dmrs_strict[order(dmrs_strict$chr, dmrs_strict$start), ]
head(dmrs_strict)

```

```{r index-dmrs}
# Define group and direction labels
test_label <- "HCT116_3BKO"
dmr_id_prefix <- paste0(test_label, "_DSS_DMR")

# Add DMR name (ID) column to dmrs
dmrs$name <- paste0(dmr_id_prefix, "_", seq_len(nrow(dmrs)), "_", toupper(dmrs$chr))

dmr_strict_id_prefix <- paste0(test_label, "_DSS_DMR_STRICT")

dmrs_strict$DMR_ID <- seq_len(nrow(dmrs_strict))
dmrs_strict$name <- paste0(dmr_strict_id_prefix, "_", dmrs_strict$DMR_ID, "_", toupper(dmrs_strict$chr))

```


```{r sort-dmrs}
# dmrs_hyper <- subset(dmrs, diff.Methy < 0)
# dmrs_hypo  <- subset(dmrs, diff.Methy > 0)

dmrs_strict_hyper <- subset(dmrs_strict, diff.Methy < 0)
dmrs_strict_hypo  <- subset(dmrs_strict, diff.Methy > 0)

dmr_strict_hyper_id_prefix <- paste0(test_label, "_DSS_DMR_STRICT_HYPER")
dmrs_strict_hypo_id_prefix <- paste0(test_label, "_DSS_DMR_STRICT_HYPO")

dmrs_strict_hyper$name <- sub("DMR_STRICT_", "DMR_STRICT_HYPER_", dmrs_strict_hyper$name)
dmrs_strict_hypo$name  <- sub("DMR_STRICT_", "DMR_STRICT_HYPO_", dmrs_strict_hypo$name)

```


## Save DMR and DML results for IGV

```{r}
options(scipen=999) # to prevent scientific notation on tables

today <- format(Sys.Date(), "%d_%m_%Y")

# Get plain file name
ref_name <- tools::file_path_sans_ext(basename(params$input1), compression = TRUE)
ref_name <- tools::file_path_sans_ext(ref_name, compression = TRUE)
test_name <- tools::file_path_sans_ext(basename(params$input2), compression = TRUE)
test_name <- tools::file_path_sans_ext(test_name, compression = TRUE)

# Create file prefix
prefix <- paste0("HG38_", today, "_", toupper(test_name), "_VS_", toupper(ref_name))

dmr_file_strict <- paste0(prefix, "_DMRs_STRICT.bed")
dmr_file <- paste0(prefix, "_DMRs.bed")
dml_file <- paste0(prefix, "_DMLs_ALL.bed")
# Additional hyp/hyper output files
dmr_file_strict_hyper <- paste0(prefix, "_DMRs_STRICT_HYPER.bed")
dmr_file_strict_hypo  <- paste0(prefix, "_DMRs_STRICT_HYPO.bed")

# Write results to files
write.table(dmrs[, c("chr", "start", "end", "name")],
            file = dmr_file,
            sep = "\t", quote = FALSE, row.names = FALSE, col.names = FALSE)

write.table(dmrs_strict[, c("chr", "start", "end", "name")],
            file = dmr_file_strict,
            sep = "\t", quote = FALSE, row.names = FALSE, col.names = FALSE)

write.table(dmls_strict[, c("chr", "pos", "pos")],
            file = dml_file,
            sep = "\t", quote = FALSE, row.names = FALSE, col.names = FALSE)

write.table(dmrs_strict_hyper[, c("chr", "start", "end", "name")],
            file = dmr_file_strict_hyper,
            sep = "\t", quote = FALSE, row.names = FALSE, col.names = FALSE)

write.table(dmrs_strict_hypo[, c("chr", "start", "end", "name")],
            file = dmr_file_strict_hypo,
            sep = "\t", quote = FALSE, row.names = FALSE, col.names = FALSE)

# Outputs
cat("Saving DMLs to:", dml_file, "\n")
cat("Saving DMRs to:", dmr_file, "\n")
cat("Saving strict DMRs to:", dmr_file_strict, "\n")

cat("Saving strict hyper DMRs to:", dmr_file_strict_hyper, "\n")
cat("Saving strict hypo DMRs to:", dmr_file_strict_hypo, "\n")
```


## Session Info

```{r session-info}
sessionInfo()
```


