---
title: "DSS Analysis"
author: "Cemile Balkas"
output: html_document
params:
  input1: "WT_WG.tidy.dssin.gz" # Reference group 
  input2: "3BKO_WG.tidy.dssin.gz" # Test group
---

To run on command line Rscript -e "rmarkdown::render('Analysis.Rmd', params = list(input1 = 'file_name', input2 = 'file_name'))"

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

This script performs analysis of differentially methylated regions (DMRs) using the DSS package. The input data is from Oxford Nanopore bisulfite sequencing and reformatted to make it compatible for DSS. DSS was chosen due to its accuracy in calling DMRs. 

## Setup

```{r itit}
# Makes sure the necessary packages are installed before running this document
if (!requireNamespace("BiocManager", quietly = TRUE)) {
  install.packages("BiocManager")
}
if (!requireNamespace("DSS", quietly = TRUE)) {
  BiocManager::install("DSS")
}

# Load required libraries
library(DSS)
require(bsseq)

```


```{r}
cat("Comparing methylation between:\n")
cat("  - Reference group:", params$input1, "\n")
cat("  - Test group:", params$input2, "\n")

```

# Read Files

```{r load-data}
# Load methylation count data: chr, pos, total reads (N), methylated reads (X)
wt_file <- params$input1
ko_file <- params$input2

wt <- read.table(wt_file, header = FALSE)
ko <- read.table(ko_file, header = FALSE)

colnames(wt) <- c("chr", "pos", "N", "X")
colnames(ko) <- c("chr", "pos", "N", "X")

# Create BSseq object
# This object holds methylation information in the structure DSS requires
BSobj <- makeBSseqData(list(wt, ko), c("WT", "KO"))

# View summary of the BSseq object
BSobj
```

## Coverage comparison across chromosomes

Before performing differential methylation analysis the read coverage between chromosomes can be checked to see if the coverage is consistent and sufficient across chromosomes. This is to make sure that the detected DMRs are not artefacts of uneven sequencing depth. 

```{r coverage-boxplots}
# Canonical chromosome levels
canon_chrs <- paste0("chr", 1:22)

# Makes the chr column is a factor with correct order
wt$chr <- factor(wt$chr, levels = canon_chrs)
ko$chr <- factor(ko$chr, levels = canon_chrs)

# Reference group coverage boxplot for each chromosome
boxplot(N ~ chr, data = wt,
        main = paste0("Coverage by Chromosome Reference Group \n(", params$input1, ")"),
        ylab = "Coverage (N)",
        xlab = "Chromosome",
        col = "skyblue",
        outline = FALSE,
        las = 2, cex.axis = 0.7)

# Test group coverage boxplot
boxplot(N ~ chr, data = ko,
        main = paste0("Coverage by Chromosome Test Group \n(", params$input2, ")"),
        ylab = "Coverage (N)",
        xlab = "Chromosome",
        col = "coral",
        outline = FALSE,
        las = 2, cex.axis = 0.7)
```
##  Differential Methylation Loci (DML) test

DML test between the two groups is performed using the DMLtest function from the DSS package. This tests each CpG site for significant differences in methylation level between groups, applying smoothing to borrow strength from neighboring sites, but is compulsary for DSS when there are no replicates. 

```{r dml-test}
# Run DML test with smoothing (recommended for WGBS data and when there are no replicates)
dmlTest.sm <- DMLtest(BSobj,  group1="WT", group2="KO", smoothing=TRUE, ncores=4)
head(dmlTest.sm)
```
## Identify DMLs based on statistical test results

With the test results, the callDML function was called. The results are DMLs are sorted by their significance.

```{r}
# Null hypothesis: the diff. in methylation levels is 0 between groups (mu1 = mu2)
# No minimum effect size required
dmls <- callDML(dmlTest.sm, p.threshold = 0.001)
head(dmls)

# Null hypothesis: the difference in methylation is less than or equal to delta (|mu1 - mu2| ≤ 0.1)
# Code below filters for biologically meaningful changes (at least 10%)
dmls_strict <- callDML(dmlTest.sm, delta = 0.1, p.threshold = 0.001)
head(dmls_strict)

# For DMLs add +2 to cover the whole CpG site as the locus
dmls_strict$end <- dmls$pos + 2

```

## Identify DMRs by grouping nearby significant CpGs (DMLs)

After the identification of differentially methylated loci (DMLs), differentially methylated regions (DMRs) are identified. The callDMR() function from DSS was used. Regions are identified based on significance thresholds for both statistical confidence (p-value) and the magnitude of methylation difference (delta). This step summarises regional methylation patterns than just the single loci differences.

```{r dmr-call}
# Detect DMRs based on smoothed DML test results
# Call DMRs using p-value threshold only (no effect size filter)
dmrs = callDMR(dmlTest.sm, p.threshold=0.01)
head(dmrs)

# Call stricter DMRs using both p-value and effect size thresholds
dmrs_strict <- callDMR(dmlTest.sm, 
                       delta = 0.1,         # minimum methylation difference (default: 0)
                       p.threshold = 0.01,  # significance threshold (default: 1e-5)
                       minlen = 50,         # minimum DMR length in bp (default: 50)
                       minCG = 3,           # minimum number of CpGs per DMR (default: 3)
                       dis.merge = 50,      # merge DMRs closer than this (default: 50 bps)
                       pct.sig = 0.9)       # % of significant CpGs required (default: 0.5)
head(dmrs_strict)

# Visualise an example DMR
showOneDMR(dmrs_strict[1,], BSobj)

# Extend end coordinate to cover the full CpG site at the end (2bp) for IGV visualisation
dmrs$end <- dmrs$end + 2
dmrs_strict$end <- dmrs_strict$end + 2

# Sort strict DMRs by chromosome and start position
# Set chromosome column as a factor with canonical order
dmrs_strict$chr <- factor(dmrs_strict$chr, levels = canon_chrs)

# Order by chromosomes (with factor level order) and then by start position
dmrs_strict <- dmrs_strict[order(dmrs_strict$chr, dmrs_strict$start), ]
head(dmrs_strict)
```
## Comparative Visualisation of DMRs

```{r}
# Function adapted from https://www.geeksforgeeks.org/outlier-analysis-in-r/
# to identify outliers using the IQR method
detect_outlier <- function(x) {

  Quantile1 <- quantile(x, probs=.25)
  Quantile3 <- quantile(x, probs=.75)
  
  IQR = Quantile3-Quantile1
  
  x > Quantile3 + (IQR*1.5) | x < Quantile1 - (IQR*1.5)
}

# Count the number of outliers in DMR length and CpG count for both sets
sum(detect_outlier(dmrs$length))
# [1] 30319
sum(detect_outlier(dmrs_strict$length))
# [1] 18626
sum(detect_outlier(dmrs$nCG))
# [1] 28528
sum(detect_outlier(dmrs_strict$nCG))
# [1] 15514

# Summary statistics to check the distributions
summary(dmrs$length)
summary(dmrs_strict$length)
summary(dmrs$nCG)
summary(dmrs_strict$nCG)

# Scatter plots: DMR length vs CpG count
par(mfrow=c(1,2))
plot(dmrs$length, dmrs$nCG, 
     xlab="Length (bp)", ylab="Number of CpG sites", 
     main="Loose DMRs:\nLength vs CpG Count", ylim = c(0, 500), xlim = c(0, 30000),
     pch=16, col=rgb(0,0,1,0.5), cex=0.7)

plot(dmrs_strict$length, dmrs_strict$nCG, 
     xlab="Length (bp)", ylab="Number of CpG sites", 
     main="Strict DMRs:\nLength vs CpG Count", ylim = c(0, 500), xlim = c(0, 30000),
     pch=16, col=rgb(1,0,0,0.5), cex=0.7)

# Boxplots to compare DMR length and CpG count distributions
boxplot(dmrs$length, dmrs_strict$length, names=c("Loose", "Strict"),
        main="DMR Lengths", ylab="Length (bp)", col=c("skyblue", "tomato"))

boxplot(dmrs$nCG, dmrs_strict$nCG, names=c("Loose", "Strict"),
        main="CpG Sites per DMR", ylab="CpG Count", col=c("lightgreen", "orange"))

# Boxplots without outliers for a clearer look at the core distribution
boxplot(dmrs$length, dmrs_strict$length, 
        names=c("Loose", "Strict"), 
        main="DMR Lengths", ylab="Length (bp)", outline=FALSE, col=c("skyblue", "tomato"))

boxplot(dmrs$nCG, dmrs_strict$nCG, 
        names=c("Loose", "Strict"), 
        main="CpG Count per DMR", ylab="# CpG sites",outline=FALSE, col=c("lightgreen", "orange"))

```
### Assign unique identifiers to each DMR

Each DMR is assigned a descriptive name to help track and label them in IGV and downstream analysis.

```{r index-dmrs}
# Define group label for use in DMR names
test_label <- "HCT116_3BKO"

# Create DMR names for the non-strict DMR set
dmr_id_prefix <- paste0(test_label, "_DSS_DMR")
dmrs$name <- paste0(dmr_id_prefix, "_", seq_len(nrow(dmrs)), "_", toupper(dmrs$chr))

# Create DMR names and numeric IDs for the strict DMR set
dmr_strict_id_prefix <- paste0(test_label, "_DSS_DMR_STRICT")
dmrs_strict$DMR_ID <- seq_len(nrow(dmrs_strict))
dmrs_strict$name <- paste0(dmr_strict_id_prefix, "_", dmrs_strict$DMR_ID, "_", toupper(dmrs_strict$chr))

```

### Split strict DMRs by methylation direction

The DMRs are split into hypermethylated and hypomethylated groups based on the direction of change in methylation between groups, and the region names are updated accordingly.

```{r sort-dmrs}
# based on the direction of methylation change (diff.Methy):
#   - Negative = hypermethylation in tested group
#   - Positive = hypomethylation in tested group

dmrs_strict_hyper <- subset(dmrs_strict, diff.Methy < 0)
dmrs_strict_hypo  <- subset(dmrs_strict, diff.Methy > 0)

dmr_strict_hyper_id_prefix <- paste0(test_label, "_DSS_DMR_STRICT_HYPER")
dmrs_strict_hypo_id_prefix <- paste0(test_label, "_DSS_DMR_STRICT_HYPO")

# Update DMR names to include methylation direction
dmrs_strict_hyper$name <- sub("DMR_STRICT_", "DMR_STRICT_HYPER_", dmrs_strict_hyper$name)
dmrs_strict_hypo$name  <- sub("DMR_STRICT_", "DMR_STRICT_HYPO_", dmrs_strict_hypo$name)

```

## Output Summary

```{r}
cat("Output Summary:\n")
cat("  Total DMLs:", nrow(dmls_strict), "\n")
cat("  Total DMRs:", nrow(dmrs_strict), "\n")
cat("    Hypermethylated:", nrow(dmrs_strict_hyper), "\n")
cat("    Hypomethylated:", nrow(dmrs_strict_hypo), "\n")

```

## Save DML and DMR results to files

"STRICT" files apply a more conservative threshold (delta = 0.1, pct.sig < 0.9, etc.) and are more likely to be actual DMRs. Regular DMRs (detected using p < 0.01 only) may include less accurate or noisy sites.

```{r}
options(scipen=999) # to prevent scientific notation on tables

today <- format(Sys.Date(), "%d_%m_%Y")

# Get plain file name
ref_name <- tools::file_path_sans_ext(basename(params$input1), compression = TRUE)
ref_name <- tools::file_path_sans_ext(ref_name, compression = TRUE)
test_name <- tools::file_path_sans_ext(basename(params$input2), compression = TRUE)
test_name <- tools::file_path_sans_ext(test_name, compression = TRUE)

# Create file prefix
prefix <- paste0("HG38_", today, "_", toupper(test_name), "_VS_", toupper(ref_name))

dmr_file_strict <- paste0(prefix, "_DMRs_STRICT.bed")
dmr_file <- paste0(prefix, "_DMRs.bed")
dml_file <- paste0(prefix, "_DMLs_ALL.bed")
# Additional hypo/hyper output files
dmr_file_strict_hyper <- paste0(prefix, "_DMRs_STRICT_HYPER.bed")
dmr_file_strict_hypo  <- paste0(prefix, "_DMRs_STRICT_HYPO.bed")

# Write results to files
write.table(dmrs[, c("chr", "start", "end", "name")],
            file = dmr_file,
            sep = "\t", quote = FALSE, row.names = FALSE, col.names = FALSE)

write.table(dmrs_strict[, c("chr", "start", "end", "name")],
            file = dmr_file_strict,
            sep = "\t", quote = FALSE, row.names = FALSE, col.names = FALSE)

write.table(dmls_strict[, c("chr", "pos", "end")],
            file = dml_file,
            sep = "\t", quote = FALSE, row.names = FALSE, col.names = FALSE)

write.table(dmrs_strict_hyper[, c("chr", "start", "end", "name")],
            file = dmr_file_strict_hyper,
            sep = "\t", quote = FALSE, row.names = FALSE, col.names = FALSE)

write.table(dmrs_strict_hypo[, c("chr", "start", "end", "name")],
            file = dmr_file_strict_hypo,
            sep = "\t", quote = FALSE, row.names = FALSE, col.names = FALSE)

# Outputs
cat("Saving DMLs to:", dml_file, "\n")
cat("Saving DMRs to:", dmr_file, "\n")
cat("Saving strict DMRs to:", dmr_file_strict, "\n")

cat("Saving strict hyper DMRs to:", dmr_file_strict_hyper, "\n")
cat("Saving strict hypo DMRs to:", dmr_file_strict_hypo, "\n")
cat("Done.")

```

## Session Info

```{r session-info}
sessionInfo()
```


