---
title: "DMR Associated Gene Expression Changes"
author: "Cemile Balkas"
date: "2025-07-08"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(readr)
library(ggplot2)
library(dplyr)
if (!requireNamespace("devel", quietly = TRUE)) {
  BiocManager::install("devel")
}
if (!requireNamespace("clusterProfiler", quietly = TRUE)) {
  BiocManager::install("clusterProfiler")
}
library(clusterProfiler)
library(org.Hs.eg.db)

library("biomartr")
library("clusterProfiler")
library("tidyverse")
library("enrichplot")
library("biomaRt")
library(ggpubr)

```

## Load DESeq2 Results

```{r load-differential-expression-table}
de <- read.table(file = "Deseq2_Genotype_DNMT3B_vs_WT_sval_annot_190623.txt", sep = "\t", header = TRUE)
nrow(de)
```

## Load DMR Gene Lists

```{r load-dmr-gene-lists}
# Define gene group files
# group of gene ids with hyper methylated tss
# group of gene ids with hyper methylated exons
# group of genes with hyoo methlated introns etc etc.
gene_list_files <- list(
  "hyper_tss" = "dmr_feature_overlaps_exclusive/hyper_tss_gene_ids.txt",
  "hyper_exons" = "dmr_feature_overlaps_exclusive/hyper_exons_gene_ids.txt",
  "hyper_introns" = "dmr_feature_overlaps_exclusive/hyper_introns_gene_ids.txt",
  "hyper_3utrs" = "dmr_feature_overlaps_exclusive/hyper_3utrs_gene_ids.txt",
  "hyper_5utrs" = "dmr_feature_overlaps_exclusive/hyper_5utrs_gene_ids.txt",
  "hyper_downstream_1kb" = "dmr_feature_overlaps_exclusive/hyper_downstream_1kb_gene_ids.txt",
  # "hyper_intergenic" = "dmr_feature_overlaps_exclusive/hyper_intergenic_gene_ids.txt",
  "hypo_tss" = "dmr_feature_overlaps_exclusive/hypo_tss_gene_ids.txt",
  "hypo_exons" = "dmr_feature_overlaps_exclusive/hypo_exons_gene_ids.txt",
  "hypo_introns" = "dmr_feature_overlaps_exclusive/hypo_introns_gene_ids.txt",
  "hypo_3utrs" = "dmr_feature_overlaps_exclusive/hypo_3utrs_gene_ids.txt",
  "hypo_5utrs" = "dmr_feature_overlaps_exclusive/hypo_5utrs_gene_ids.txt",
  "hypo_downstream_1kb" = "dmr_feature_overlaps_exclusive/hypo_downstream_1kb_gene_ids.txt"
  # "hypo_intergenic" = "dmr_feature_overlaps_exclusive/hypo_intergenic_gene_ids.txt"
)

# Load into a named list of dataframes
gene_groups <- lapply(gene_list_files, read_lines)
```

## Compare Log2FC Distributions (without p values)

```{r log2fc-distributions-simple}
# combine all gene lists (tss, introns etc) into one big table with labels
plot_df <- bind_rows(
  lapply(names(gene_groups), function(group_name) {
    tibble(Ensembl_id = gene_groups[[group_name]], group = group_name)
  })
) %>%
  left_join(de, by = "Ensembl_id") %>% # Join with DE results (log2FC etc.) on Ensembl gene ID
  filter(!is.na(log2FoldChange.y)) # Keep only genes that are in the DESeq2 results

# define the order for the plots
hyper_features <- c(
  "hyper_tss", "hyper_5utrs", "hyper_exons", 
  "hyper_introns", "hyper_3utrs", "hyper_downstream_1kb"
)

hypo_features  <- c(
  "hypo_tss", "hypo_5utrs", "hypo_exons",
  "hypo_introns", "hypo_3utrs", "hypo_downstream_1kb")

# Filter the plot_df for only the genes that are in each the group
plot_df_hyper <- plot_df %>% filter(group %in% hyper_features)
plot_df_hypo <- plot_df %>% filter(group %in% hypo_features)


# Set factor levels to change the order of the features in plots
plot_df_hyper <- plot_df_hyper %>%
  mutate(group = factor(group, levels = hyper_features))
plot_df_hypo  <- plot_df_hypo %>%
  mutate(group = factor(group, levels = hypo_features))

custom_colors <- c(
  "hyper_tss" = "#c00606", "hypo_tss" = "#c00606",
  "hyper_5utrs" = "#ffdb00", "hypo_5utrs" = "#ffdb00",
  "hyper_exons" = "#49ff00", "hypo_exons" = "#49ff00",
  "hyper_introns" = "#00ff92", "hypo_introns" = "#00ff92",
  "hyper_3utrs" = "#0092ff", "hypo_3utrs" = "#0092ff",
  "hyper_downstream_1kb" = "#3406ab", "hypo_downstream_1kb" = "#3406ab"
)

# Hyper plot
ggplot(plot_df_hyper, aes(x = group, y = log2FoldChange.y, fill = group)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.8) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray40") +
  coord_cartesian(ylim = c(-8, 8)) +
  scale_fill_manual(values = custom_colors) +
  labs(
    title = "log2FoldChange of Hypermethylated DMR Groups",
    x = "DMR Feature Group",
    y = "log2FoldChange (3BKO vs WT)"
  ) +
  theme_classic() +
  theme(
    plot.title = element_text(colour = "#CD2626"),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 9),
    legend.position = "none"
  ) 
# Hypo plot
ggplot(plot_df_hypo, aes(x = group, y = log2FoldChange.y, fill = group)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.8) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray40") +
  coord_cartesian(ylim = c(-8, 8)) +
  scale_fill_manual(values = custom_colors) +
  labs(
    title = "log2FoldChange of Hypomethylated DMR Groups",
    x = "DMR Feature Group",
    y = "log2FoldChange (3BKO vs WT)"
  ) +
  theme_classic() +
  theme(
    plot.title = element_text(color = "#3A5FCD"),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 9),
    legend.position = "none"
  )

```
From this plot it can be seen that the hyper_tss group shows negative median log2FC suggesting that genes with hypermethylated TSSs are often down regulated in the KO dataset. (This is consistent with what we expected to see as it is widely accepted that mehtylation in promoters represses transcription)

5' UTRs being closer to the transcription start site their overall expression levels seem to have gone down, but not as greately as the hyper_tss group. 

However there is only slight increase in expression levels in the hypo_tss group, could this mean loss in methylation at TSS of genes does not necesarrily increase the expression levels of those genes???

Rest of the groups hover around 0 not showing any extreme biases in changes in expression levels. The variance in most box plots also seem to tight except the hyper_tss group.


# Wilcoxon rank-sum tests

Are the log2FoldChanges for genes with, hypermethylated TSSs systematically different from other genes?

```{r rank-sum-test-sinlge group}

# Compare log2FCs of "hyper_tss" to all other DMR-associated genes
group <- "hyper_tss"

group_values <- plot_df$log2FoldChange.y[plot_df$group == group]
# rest_values  <- plot_df$log2FoldChange.y[plot_df$group != group]

test <- wilcox.test(group_values, de_autosomal_protein_coding$log2FoldChange.y, alternative = "two.sided")

test

```


```{r wilcoxon-rank-sum-test-all}
# Filter DESeq2 results to only protein-coding genes and cannonical autosomes for the test as a background
autosomes <- as.character(1:22)

de_autosomal_protein_coding <- de %>%
  filter(gene_biotype == "protein_coding" & chromosome_name %in% autosomes)

# df for test results
wilcox_results <- lapply(unique(plot_df$group), function(g) {
  group_values <- plot_df$log2FoldChange.y[plot_df$group == g]
  background_values  <- de_autosomal_protein_coding$log2FoldChange.y

  test <- wilcox.test(group_values, background_values, alternative = "two.sided")

  tibble(
    group = g,
    p_value = test$p.value,
    median_log2FC = median(group_values),
    n = length(group_values)
  )
}) %>% bind_rows()

wilcox_results <- wilcox_results %>%
  mutate(p_adj = p.adjust(p_value, method = "BH"))

wilcox_results %>%
  arrange(p_adj) %>%
  mutate(significance = case_when(
    # p_adj <= 0.0001 ~ "****",
    p_adj <= 0.001 ~ "***",
    p_adj <= 0.01  ~ "**",
    p_adj <= 0.05  ~ "*",
    TRUE          ~ "ns"
  ))

```
## Compare Log2FC Distributions (with p values)

```{r fold-change-plots with p values}
autosomes <- as.character(1:22)

de_autosomal_protein_coding <- de %>%
  filter(gene_biotype == "protein_coding" & chromosome_name %in% autosomes)

background_df <- de_autosomal_protein_coding %>%
  mutate(group = "background") %>%
  dplyr::select(Ensembl_id, log2FoldChange.y, group)

# subset the data frames to columns needed
plot_df_hypo_subset <- plot_df %>%
  filter(group %in% hypo_features) %>%
  dplyr::select(Ensembl_id, log2FoldChange.y, group)

plot_df_hyper_subset <- plot_df %>%
  filter(group %in% hyper_features) %>%
  dplyr::select(Ensembl_id, log2FoldChange.y, group)

combined_df_hypo <- bind_rows(plot_df_hypo_subset, background_df)
combined_df_hyper <- bind_rows(plot_df_hyper_subset, background_df)

# Set group order for plotting
combined_df_hypo$group <- factor(combined_df_hypo$group, levels = c("background", hypo_features))
combined_df_hyper$group <- factor(combined_df_hyper$group, levels = c("background", hyper_features))

# Plot for hyper features -----------------------------------------------------
pvals_df <- lapply(hyper_features, function(g) {
  test <- wilcox.test(
    combined_df_hyper$log2FoldChange.y[combined_df_hyper$group == g],
    combined_df_hyper$log2FoldChange.y[combined_df_hyper$group == "background"]
  )
  data.frame(group1 = "background", group2 = g, p = test$p.value)
}) %>% bind_rows()

pvals_df <- pvals_df %>%
  mutate(
    y.position = 6,
    p.label = ifelse(p < 2e-16, "p = <2e-16", sprintf("p = %.3g", p))
  )

ggboxplot(
  data = combined_df_hyper %>% filter(group != "background"),
  x = "group", y = "log2FoldChange.y", fill = "group", outlier.shape = NA
) +
  scale_fill_manual(values = custom_colors) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray40") +
  coord_cartesian(ylim = c(-8, 8)) +
  labs(
    title = "log2FoldChange of Hypermethylated DMR Groups",
    x = "DMR Feature Group",
    y = "log2FoldChange (3BKO vs WT)"
  ) +
  theme_classic() +
  theme(
    plot.title = element_text(colour = "#CD2626"),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 9),
    plot.margin = margin(t = 10, r = 10, b = 10, l = 10),
    legend.position = "none"
  ) +
  stat_pvalue_manual(pvals_df, label = "p.label", xmin = "group2", xmax = "group2", size = 3.6)
# ggsave("analysis_results/gene_log2foldchange_hyper_dmr.png", height = 6.5, units = "in")

# Plot for hypo features --------------------------------------------------------------
pvals_df <- lapply(hypo_features, function(g) {
  test <- wilcox.test(
    combined_df_hypo$log2FoldChange.y[combined_df_hypo$group == g],
    combined_df_hypo$log2FoldChange.y[combined_df_hypo$group == "background"]
  )
  data.frame(group1 = "background", group2 = g, p = test$p.value)
}) %>% bind_rows()

pvals_df <- pvals_df %>%
  mutate(
    y.position = 6,
    p.label = ifelse(p < 2e-16, "p = <2e-16", sprintf("p = %.3g", p))
  )

ggboxplot(
  data = combined_df_hypo %>% filter(group != "background"),
  x = "group", y = "log2FoldChange.y", fill = "group", outlier.shape = NA
) +
  scale_fill_manual(values = custom_colors) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray40") +
  coord_cartesian(ylim = c(-8, 8)) +
  labs(
    title = "log2FoldChange of Hypomethylated DMR Groups",
    x = "DMR Feature Group",
    y = "log2FoldChange (3BKO vs WT)"
  ) +
  theme_classic() +
  theme(
    plot.title = element_text(color = "#3A5FCD"),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 9),
    plot.margin = margin(t = 10, r = 10, b = 10, l = 10),
    legend.position = "none"
  ) +
  stat_pvalue_manual(pvals_df, label = "p.label", xmin = "group2", xmax = "group2", size = 3.6)
# ggsave("analysis_results/gene_log2foldchange_hypo_dmr.png", height = 6.5, units = "in")

```

## How many genes are there per group? 
## How many are up/down regulated 

```{r count-genes-per-group}
# Function that returns a data frame summarising total, up, and down-regulated gene counts per DMR group
summarise_dmr_expression_counts <- function(group_names, gene_groups, de_data, log2fc_cutoff = 1, padj_cutoff = 0.05) {
  summary_list <- lapply(group_names, function(group_name) {
    # Get the Ensembl gene IDs for this group
    ids <- gene_groups[[group_name]]

    # Subset the DESeq2 data to just these genes
    sub_de <- subset(de_data, Ensembl_id %in% ids)

    # Identify significantly differentially expressed genes
    sig <- subset(sub_de, padj < padj_cutoff & abs(log2FoldChange.y) > log2fc_cutoff)

    # Count up and down regulated genes
    up <- sum(sig$log2FoldChange.y > log2fc_cutoff, na.rm = TRUE)
    down <- sum(sig$log2FoldChange.y < -log2fc_cutoff, na.rm = TRUE)
    total <- nrow(sub_de)

    data.frame(
      Group = group_name,
      Total = total,
      Upregulated = up,
      Downregulated = down
    )
  })

  # Combine individual rows into a single summary data frame
  do.call(rbind, summary_list)
}

# Create a summary table of gene expression changes for each DMR group
dmr_gene_expression_summary <- summarise_dmr_expression_counts(
  group_names = names(gene_groups),
  gene_groups = gene_groups,
  de_data = de
)

print(dmr_gene_expression_summary)
```
Here the genes were grouped based on where differentially methylated regions (DMRs) intersected genomic features. Since a single gene can have multiple associated DMRs across different features, individual genes may appear in more than one group (eg a gene with both a hypermethylated TSS and hypermethylated exons will be represented in both groups). 

The summary table reflects feature specific patterns of expression change rather than a count of unique genes. 

This does not affect the interpretation of each group independently, but total numbers of up- or downregulated genes across multiple DMR categories shouldn't be generalised without deduplication.

## Define funtion to get up/down regulated genes by group 

```{r get up/down regulated genes}
get_up_down_genes <- function(group_name, log2fc_cutoff = 1, padj_cutoff = 0.05) {
  # Get Ensembl IDs
  ids <- gene_groups[[group_name]]

  # Filter DE results
  sub_de <- subset(de, Ensembl_id %in% ids)
  sig <- subset(sub_de, padj < padj_cutoff & abs(log2FoldChange.y) > log2fc_cutoff)

  # Separate up and down and order by padjust value
  up_genes <- sig %>%
    filter(log2FoldChange.y > log2fc_cutoff) %>%
    arrange(padj)

  down_genes <- sig %>%
    filter(log2FoldChange.y < -log2fc_cutoff) %>%
    arrange(padj)

  list(
    up = up_genes %>% dplyr::select(Ensembl_id, external_gene_name, log2FoldChange.y, padj),
    down = down_genes %>% dplyr::select(Ensembl_id, external_gene_name, log2FoldChange.y, padj)
  )
}
```

## Check genes in groups of interest

```{r}
# List of DMR groups I'd like to look into
features_to_check <- c("hyper_exons", "hypo_exons", "hyper_tss", "hypo_tss", "hyper_introns", "hypo_introns", "hyper_5utrs")

# Apply the get_up_down_genes function to each group in the list,
# and return a named list where each entry corresponds to that group's results
dmr_gene_expression <- setNames(
  lapply(features_to_check, function(group) {
    # For each group name get up/downregulated genes
    get_up_down_genes(group)
  }),
  features_to_check
)
```

# ORA with clusterProfiler

The background gene set for over-representation analysis was limited to protein-coding genes. This is because only protein-coding genes were used when linking DMRs to genes, and all other regions were treated as intergenic. Using all genes from the DESeq2 results, including non-coding RNAs, would not match the gene sets used in the analysis. To keep things consistent, only protein-coding genes of cannonical autosomal crhommosomes from the DESeq2 output were used as the universe in the enrichment analysis.

```{r buiding-the-universe}
# building the universe!

# Filter DESeq2 results to only protein-coding genes and cannonical autosomes
autosomes <- as.character(1:22)

de_autosomal_protein_coding <- de %>%
  filter(gene_biotype == "protein_coding" & chromosome_name %in% autosomes)

# Create universe from filtered DE table
universe_genes <- unique(de_autosomal_protein_coding$Ensembl_id)


# If just protein coding uncomment this code:
# de_protein_coding <- subset(de, gene_biotype == "protein_coding")
#
# # Get universe: all protein-coding genes tested in RNA-seq
# universe_genes <- unique(de_protein_coding$Ensembl_id)
```

## Hyper-TSS & Hyper-5UTR/ Downregulated genes

What are the genes that downregulated with hyper methylated transcription start sites? and are they enriched in any biological pathways, cellular components or molecular functions?

Genes with hypermethylated transcription start sites (TSSs) and 5' untranslated regions (5'UTRs) that were significantly downregulated were tested for functional enrichment using Gene Ontology (GO) analysis. No significant enrichment was detected for either group for Biological Process (BP), Cellular Component (CC), or Molecular Function (MF) categories, even with a less stringent q-value cutoff. While hypermethylation at TSSs is generally associated with gene silencing, this analysis did not identify any specific biological themes among the downregulated genes. 

This could mean functional heterogeneity among affected genes, or subtle effects not captured by over representation analysis.

```{r}
# Top 10 genes that are most significantly down-regulated with hyper_tss
dmr_gene_expression$hyper_tss$down$external_gene_name %>% head(n = 10)

ora_hyper_tss_down <- enrichGO(
  gene          = dmr_gene_expression$hyper_tss$down$Ensembl_id,
  universe      = universe_genes,
  OrgDb         = org.Hs.eg.db,
  keyType       = "ENSEMBL",
  ont           = "BP",
  pAdjustMethod = "BH",
  qvalueCutoff  = 0.1,
  readable      = TRUE,
  pool          = FALSE
)

# Simplify and check results
ora_hyper_tss_down_simplified <- clusterProfiler::simplify(ora_hyper_tss_down)
head(ora_hyper_tss_down_simplified@result)

# Only plot if there are results
if (nrow(ora_hyper_tss_down_simplified@result) > 0) {
  dotplot(ora_hyper_tss_down_simplified)
} else {
  message("No significant enrichment results to plot.")
}

# Top 10 genes that are most significantly down-regulated with hyper_5utrs
dmr_gene_expression$hyper_5utrs$down$external_gene_name %>% head(n = 10)

ora_hyper_5utrs_down <- enrichGO(
  gene          = dmr_gene_expression$hyper_5utrs$down$Ensembl_id,
  universe      = universe_genes,
  OrgDb         = org.Hs.eg.db,
  keyType       = "ENSEMBL",
  ont           = "CC",
  pAdjustMethod = "BH",
  qvalueCutoff  = 0.1,
  readable      = TRUE,
  pool          = FALSE
)

# Simplify and check results
ora_hyper_5utrs_down_simplified <- clusterProfiler::simplify(ora_hyper_5utrs_down)
head(ora_hyper_5utrs_down_simplified@result)

# Only plot if there are results
if (nrow(ora_hyper_5utrs_down_simplified@result) > 0) {
  dotplot(ora_hyper_5utrs_down_simplified)
} else {
  message("No significant enrichment results to plot.")
}
```

## Hyper-Exons / Downregulated genes

For downregulated genes associated with hypermethylated exons, no significant enrichment was found in the Molecular Function (MF) or Cellular Component (CC) categories. In the Biological Process (BP) category, enrichment was observed for terms related to extracellular matrix organisation & epithelial cell proliferation. 

This could mean that hypermethylation in exons may be linked to the repression of genes involved in structural tissue organisation and cell growth??? 

```{r}
# Top 10 genes that are most significantly down-regulated with hyper_tss
dmr_gene_expression$hyper_exons$down$external_gene_name %>% head(n = 10)

ora_hyper_exons_down <- enrichGO(
  gene          = dmr_gene_expression$hyper_exons$down$Ensembl_id,
  universe      = universe_genes,
  OrgDb         = org.Hs.eg.db,
  keyType       = "ENSEMBL",
  ont           = "BP",
  pAdjustMethod = "BH",
  qvalueCutoff  = 0.05,
  readable      = TRUE,
  pool          = FALSE
)

# Simplify and check results
ora_hyper_exons_down_simplified <- clusterProfiler::simplify(ora_hyper_exons_down)
head(ora_hyper_exons_down_simplified@result)

# Only plot if there are results
if (nrow(ora_hyper_exons_down_simplified@result) > 0) {
  dotplot(ora_hyper_exons_down_simplified)
} else {
  message("No significant enrichment results to plot.")
}
```

## Hyper-Introns / Downregulated genes

Downregulated genes associated with hypermethylated introns showed enrichment for multiple biological processes (BP), mainly related to cell proliferation, migration, and angiogenesis. No enrichment was found for molecular function (MF), but enrichment was observed in the cellular component (CC) ontology, including terms such as extracellular matrix, focal adhesion, and membrane related components.

This suggests intronic hypermethylation is linked to transcriptional repression of genes involved in tissue structure and cell signalling and movement.


```{r}
# Top 10 genes that are most significantly down-regulated with hyper_tss
dmr_gene_expression$hyper_introns$down$external_gene_name %>% head(n = 10)

ora_hyper_introns_down_bp <- enrichGO(
  gene          = dmr_gene_expression$hyper_introns$down$Ensembl_id,
  universe      = universe_genes,
  OrgDb         = org.Hs.eg.db,
  keyType       = "ENSEMBL",
  ont           = "BP",
  pAdjustMethod = "BH",
  qvalueCutoff  = 0.05,
  readable      = TRUE,
  pool          = FALSE
)

# Simplify and check results
ora_hyper_introns_down_simplified_bp <- clusterProfiler::simplify(ora_hyper_introns_down_bp)
head(ora_hyper_introns_down_simplified@result)

# Only plot if there are results
if (nrow(ora_hyper_introns_down_simplified_bp@result) > 0) {
  dotplot(ora_hyper_introns_down_simplified_bp)
} else {
  message("No significant enrichment results to plot.")
}

ora_hyper_introns_down_simplified_bp <- pairwise_termsim(ora_hyper_introns_down_simplified_bp, method = "JC")
emapplot(ora_hyper_introns_down_simplified_bp, showCategory = 15, color = "qvalue")

# Molecular Function
ora_hyper_introns_down_cc <- enrichGO(
  gene          = dmr_gene_expression$hyper_introns$down$Ensembl_id,
  universe      = universe_genes,
  OrgDb         = org.Hs.eg.db,
  keyType       = "ENSEMBL",
  ont           = "CC",
  pAdjustMethod = "BH",
  qvalueCutoff  = 0.05,
  readable      = TRUE,
  pool          = FALSE
)

# Simplify and check results
ora_hyper_introns_down_simplified_cc <- clusterProfiler::simplify(ora_hyper_introns_down_cc)
head(ora_hyper_introns_down_simplified_cc@result)

# Only plot if there are results
if (nrow(ora_hyper_introns_down_simplified_cc@result) > 0) {
  dotplot(ora_hyper_introns_down_simplified_cc)
} else {
  message("No significant enrichment results to plot.")
}

```

## Downregulated genes that have both hyper_exons and hyper_introns

A subset of genes that are downregulated and have hypermethylation in both exonic and intronic regions. GO enrichment analysis of these genes showed significant associations with 3 biological processes:

- Extracellular matrix organisation

- Epithelial cell proliferation

- Muscle cell differentiation 

No significant enrichment was observed for molecular function (MF) or cellular component (CC) ontologies.


```{r}
# Find any genes that are down regulated and have both hypermethylated exons and introns
downregulated_in_exon_and_intron <- intersect(
  dmr_gene_expression$hyper_exons$down$Ensembl_id,
  dmr_gene_expression$hyper_introns$down$Ensembl_id
)

# Subset the DE table to get metadata for these genes
overlap_gene_info <- de %>%
  filter(Ensembl_id %in% downregulated_in_exon_and_intron) %>%
  dplyr::select(Ensembl_id, external_gene_name, log2FoldChange.y, padj, gene_biotype, chromosome_name)

# Perform GO enrichment
ora_overlap_downregulated_bp <- enrichGO(
  gene         = overlap_gene_info$Ensembl_id,
  universe     = universe_genes,
  OrgDb        = org.Hs.eg.db,
  keyType      = "ENSEMBL",
  ont          = "BP",
  pAdjustMethod = "BH",
  qvalueCutoff  = 0.05,
  readable      = TRUE
)

ora_overlap_downregulated_simplified_bp <- clusterProfiler::simplify(ora_overlap_downregulated_bp)
head(ora_overlap_downregulated_simplified_bp@result)

# Visualise enrichment
if (nrow(ora_overlap_downregulated_simplified_bp@result) > 0) {
  dotplot(ora_overlap_downregulated_simplified_bp)
} else {
  message("No significant enrichment results to plot.")
}

ora_overlap_downregulated_bp <- pairwise_termsim(ora_overlap_downregulated_bp, method = "JC")
emapplot(ora_overlap_downregulated_bp, color = "qvalue")

# # Molecular Function 
# ora_overlap_downregulated_mf <- enrichGO(
#   gene          = overlap_gene_info$Ensembl_id,
#   universe      = universe_genes,
#   OrgDb         = org.Hs.eg.db,
#   keyType       = "ENSEMBL",
#   ont           = "MF",
#   pAdjustMethod = "BH",
#   qvalueCutoff  = 0.05,
#   readable      = TRUE,
#   pool          = FALSE
# )
# 
# ora_overlap_downregulated_simplified_mf <- clusterProfiler::simplify(ora_overlap_downregulated_mf)
# head(ora_overlap_downregulated_simplified_mf@result)
# 
# if (nrow(ora_overlap_downregulated_simplified_mf@result) > 0) {
#   dotplot(ora_overlap_downregulated_simplified_mf)
# } else {
#   message("No significant enrichment results to plot.")
# }
```



