---
title: "final_dss_analysis"
author: "Cemile Balkas"
date: "2025-06-18"
output: html_document
params:
  input1: "data/HCT116_WT_tidy.dssin.gz" # Reference group 
  input2: "data/HCT116_3BKO_tidy.dssin.gz" # Test group
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# Makes sure the necessary packages are installed before running this document
if (!requireNamespace("BiocManager", quietly = TRUE)) {
  install.packages("BiocManager")
}
if (!requireNamespace("DSS", quietly = TRUE)) {
  BiocManager::install("DSS")
}

# Load required libraries
library(DSS)
require(bsseq)
```

## Introduction

This script performs analysis of differentially methylated regions (DMRs) using the DSS package. The input data is from Oxford Nanopore bisulfite sequencing and reformatted to make it compatible for DSS. DSS was chosen due to its accuracy in calling DMRs. 

```{r}
cat("Comparing methylation between:\n")
cat("  - Reference group:", params$input1, "\n")
cat("  - Test group:", params$input2, "\n")

```

# Read Files

```{r load-data}
# Load methylation count data: chr, pos, total reads (N), methylated reads (X)
ref_file <- params$input1
test_file <- params$input2

ref <- read.table(ref_file, header = FALSE)
test <- read.table(test_file, header = FALSE)

colnames(ref) <- c("chr", "pos", "N", "X")
colnames(test) <- c("chr", "pos", "N", "X")

# Create BSseq object
# This object holds methylation information in the structure DSS requires
BSobj <- makeBSseqData(list(ref, test), c("REF", "TEST"))

# View summary of the BSseq object
BSobj
```

##  Differential Methylation Loci (DML) test

DML test between the two groups is performed using the DMLtest function from the DSS package. This tests each CpG site for significant differences in methylation level between groups, applying smoothing to borrow strength from neighboring sites, but is compulsary for DSS when there are no replicates. 

```{r dml-test}
# Run DML test with smoothing (recommended for WGBS data and when there are no replicates)
dmlTest.sm <- DMLtest(BSobj,  group1="REF", group2="TEST", smoothing=TRUE, ncores=4)

cat("Total mehtylated loci before calling DMLs:", nrow(dmlTest.sm), "\n")
head(dmlTest.sm)
dmlTest.sm$end <- dmlTest.sm$pos + 2

```

## Identify DMLs based on statistical test results

With the test results, the callDML function was called. The results are DMLs are sorted by their significance.

```{r}
options(scipen = 0)

# Null hypothesis: the difference in methylation is less than or equal to delta (|mu1 - mu2| ≤ 0.1)
# Code below filters for biologically meaningful changes (at least 10%)
dmls <- callDML(dmlTest.sm, delta = 0.1, p.threshold = 0.01)
head(dmls)

# For DMLs add +2 to cover the whole CpG site as the locus
dmls$end <- dmls$pos + 2

canon_chrs <- paste0("chr", 1:22)

dmls$chr <- factor(dmls$chr, levels = canon_chrs)
# Order by chromosomes (with factor level order) and then by start position
dmls <- dmls[order(dmls$chr, dmls$pos), ]
```

## Identify DMRs by grouping nearby significant CpGs (DMLs)

After the identification of differentially methylated loci (DMLs), differentially methylated regions (DMRs) are identified. The callDMR() function from DSS was used. Regions are identified based on significance thresholds for both statistical confidence (p-value) and the magnitude of methylation difference (delta). This step summarises regional methylation patterns than just the single loci differences.

```{r dmr-call}
options(scipen = 0)
# Detect DMRs based on smoothed DML test results
# Call DMRs using p-value threshold only (no effect size filter)
dmrs = callDMR(dmlTest.sm, delta = 0.1, p.threshold=0.01)
head(dmrs)

showOneDMR(dmrs_strict[1,], BSobj)

# Extend end coordinate to cover the full CpG site at the end (2bp) for IGV visualisation
dmrs$end <- dmrs$end + 2

# Sort DMRs by chromosome and start position
# Set chromosome column as a factor with canonical order
canon_chrs <- paste0("chr", 1:22)
dmrs$chr <- factor(dmrs$chr, levels = canon_chrs)
# Order by chromosomes (with factor level order) and then by start position
dmrs <- dmrs[order(dmrs$chr, dmrs$start), ]
```

### Assign unique identifiers to each DMR

Each DMR is assigned a descriptive name to help track and label them in IGV and downstream analysis.

```{r index-dmrs}
# Get plain input file names
ref_name <- tools::file_path_sans_ext(basename(params$input1), compression = TRUE)
test_name <- tools::file_path_sans_ext(basename(params$input2), compression = TRUE)

# Define group label for use in DMR names
test_label <- paste0(test_name)

# Create DMR names
dmr_id_prefix <- paste0(test_label, "_DSS_DMR")
dmrs$DMR_ID <- seq_len(nrow(dmrs))
dmrs$name <- paste0(dmr_id_prefix, "_", dmrs$DMR_ID, "_", toupper(dmrs$chr))

```

## Split DMLs by methylation direction

```{r}
dml_id_prefix <- paste0(test_label, "_DSS_DML")
dmls$DML_ID <- seq_len(nrow(dmls))
dmls$name <- paste0(dml_id_prefix, "_", dmls$DML_ID, "_", toupper(dmls$chr))

dmls_hyper <- subset(dmls, diff < 0)
dmls_hypo <- subset(dmls, diff > 0)

dmls_hyper_id_prefix <- paste0(test_label, "_DSS_DML_HYPER")
dmls_hypo_id_prefix <- paste0(test_label, "_DSS_DML_HYPO")

dmls_hyper$name <- sub("DSS_", "_DML_HYPER_", dmls_hyper$name)
dmls_hypo$name  <- sub("DSS_", "_DML_HYPO_", dmls_hypo$name)

```

### Split DMRs by methylation direction

The DMRs are split into hypermethylated and hypomethylated groups based on the direction of change in methylation between groups, and the region names are updated accordingly.

```{r sort-dmrs}
# based on the direction of methylation change (diff.Methy):
#   - Negative = hypermethylation in tested group
#   - Positive = hypomethylation in tested group

dmrs_hyper <- subset(dmrs, diff.Methy < 0)
dmrs_hypo <- subset(dmrs, diff.Methy > 0)

dmrs_hyper_id_prefix <- paste0(test_label, "_DSS_DMR_HYPER")
dmrs_hypo_id_prefix <- paste0(test_label, "_DSS_DMR_HYPO")

dmrs_hyper$name <- sub("DMR_", "DMR_HYPER_", dmrs_hyper$name)
dmrs_hypo$name  <- sub("DMR_", "DMR_HYPO_", dmrs_hypo$name)

```

## Output Summary

```{r}
cat("Output Summary:\n")
cat("  Total methylated Loci:", nrow(dmlTest.sm), "\n")
cat("  Total DMLs:", nrow(dmls), "\n")
cat("    Hypermethylated:", nrow(dmls_hyper), "\n")
cat("    Hypomethylated:", nrow(dmls_hypo), "\n")
cat("  Total DMRs:", nrow(dmrs), "\n")
cat("    Hypermethylated:", nrow(dmrs_hyper), "\n")
cat("    Hypomethylated:", nrow(dmrs_hypo), "\n")
```

## Save DML and DMR results to files

"STRICT" files apply a more conservative threshold (delta = 0.1, p < 0.05, etc.) and are more likely to be actual DMRs. Regular DMRs (detected using p < 0.01 only) may include less accurate or noisy sites.

```{r}
options(scipen=999) # to prevent scientific notation on tables

today <- format(Sys.Date(), "%d_%m_%Y")

# Create file prefix
prefix <- paste0("HG38_", today, "_", toupper(test_name), "_VS_", toupper(ref_name))

all_loci_file <- paste0(prefix, "_CpG_Loci.bed")
dml_file <- paste0(prefix, "_DMLs.bed")
# Additional hypo/hyper output files
dml_file_hyper <- paste0(prefix, "_DMLs_HYPER.bed")
dml_file_hypo  <- paste0(prefix, "_DMLs_HYPO.bed")


dmr_file <- paste0(prefix, "_DMRs.bed")
# Additional hypo/hyper output files
dmr_file_hyper <- paste0(prefix, "_DMRs_HYPER.bed")
dmr_file_hypo  <- paste0(prefix, "_DMRs_HYPO.bed")

# Write results to files
write.table(dmlTest.sm[, c("chr", "pos", "end")],
            file = all_loci_file,
            sep = "\t", quote = FALSE, row.names = FALSE, col.names = FALSE)

write.table(dmls[, c("chr", "pos", "end", "name")],
            file = dml_file,
            sep = "\t", quote = FALSE, row.names = FALSE, col.names = FALSE)

write.table(dmls_hyper[, c("chr", "pos", "end", "name")],
            file = dml_file_hyper,
            sep = "\t", quote = FALSE, row.names = FALSE, col.names = FALSE)

write.table(dmls_hypo[, c("chr", "pos", "end", "name")],
            file = dml_file_hypo,
            sep = "\t", quote = FALSE, row.names = FALSE, col.names = FALSE)

write.table(dmrs[, c("chr", "start", "end", "name")],
            file = dmr_file,
            sep = "\t", quote = FALSE, row.names = FALSE, col.names = FALSE)

write.table(dmrs_hyper[, c("chr", "start", "end", "name")],
            file = dmr_file_hyper,
            sep = "\t", quote = FALSE, row.names = FALSE, col.names = FALSE)

write.table(dmrs_hypo[, c("chr", "start", "end", "name")],
            file = dmr_file_hypo,
            sep = "\t", quote = FALSE, row.names = FALSE, col.names = FALSE)

# Outputs
cat("Saving all methylated loci to:", all_loci_file, "\n")
cat("Saving DMLs to:", dml_file, "\n")
cat("Saving hyper DMLs to:", dml_file_hyper, "\n")
cat("Saving hypo DMLs to:", dml_file_hypo, "\n")
cat("Saving DMRs to:", dmr_file, "\n")
cat("Saving hyper DMRs to:", dmr_file_hyper, "\n")
cat("Saving hypo DMRs to:", dmr_file_hypo, "\n")
cat("Done.")

```

## Session Info

```{r session-info}
sessionInfo()
```


